The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-12 manuscript/manuscript.Rnw
discuss LGALS13

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="LGALS13" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{\genetreeexplanation{galectins}}
\label{fig:galectins_lgals13_evolution_tree}
\end{figure}

Galectins are recognized as major regulators of pregnancy in both
fetal and maternal tissues (Than et al., 2012). Indeed, the evolution
of new lineage-specific galectins has occurred in both primates and
ruminants, with the genes LGALS13, LGALS14, and LGALS16 recognized as
being placenta-specific in humans and regulating immune responses by
inducing apoptosis of maternal T-cells (Than et al., 2009). Indeed,
LGALS14 has the highest expression of any galectin in the human
placenta (Table 8), and other primate galectins such as LGALS13 and
LGALS16 are also expressed in human placenta. However, in the closely
related bonobo, LGALS14 is not found, and LGALS1 is the most abundant
galectin. LGALS1 shows generally high expression in all placentas
analyzed. LGALS1 is also involved in mediating maternal-fetal immune
tolerance, and is expressed in both maternal and fetal tissue in
primates (Than et al., 2008). LGALS1 placental expression was thought
to be confined to eutherian mammals, as there is evidence of adaptive
evolution at this node, as well as the appearance of estrogen response
elements in eutherian cis-regulatory regions (Than et al., 2008). It
was unclear as to whether galectins such as LGALS1 were important in
the regulation of pregnancy in marsupials, as well as eutherians. Here
we show substantial expression of galectins in the opossum, especially
LGALS1 (Table 8). In addition, LGALS2, LGALS3, and LGALS8 are also
expressed. The presence of galectins at the maternal-fetal interface
in the opossum implies that it was also present in the ancestral
placenta of therian mammals. Indeed, adaptive evolution of LGALS1 on
the eutherian stem is likely due to the increased invasive nature of
eutehrian placentation rather than the origination of maternal-fetal
tolerance. Although the function of LGALS1 in the marsupial placenta
needs to be confirmed, it is likely that the mediation of
maternal-fetal immune tolerance by galectins is an ancient feature of
mammalian viviparity.

@@

The placenta of the opossum Monodelphis domestica is distinctly
different from that of the term placenta of eutherian mammals in that
it is derived from both the chorion and yolk sac. Indeed other
differences between the placentas of marsupials and eutherians include
lack of invasive extravillous trophoblasts in marsupials, the
retention of a permeable shell coat for 2/3 of marsupial gestation,
and the greater degree of imprinting in placenta-expressed genes in
eutherians (Renfree, 2010; Renfree et al., 2013). In addition, the
placenta of Monodelphis is also distinct from other marsupials, in
that it is considered one of the least specialized, and thought be
closer to the original marsupial condition (Renfree, 2010). In
Monodelphis, the duration of direct connection between maternal and
fetal tissues is as short as two days in a gestational period of 13-15
days. During this period, the placenta is directly engaged in the
exchange of urea with maternal tissues, owing to the lack of a
vesicular allantois (Freyer et al., 2007). This differs drastically
with the situation in another well-studied marsupial, the tammar
wallaby (Freyer et al., 2007). Therefore, we must interpret
differences between the opossum and eutherian placentas with caution,
as highly expressed genes and functional annotation clusters may
represent didelphid or Monodelphis specific adaptations.

Genes that show a significant shift on the opossum lineage include
functional annotation clusters that correspond to active generation of
energy, transport, and lipid breakdown via the peroxisome (Table 10;
Supplemental Table X). The choriovitelline placenta of both didelphids
and the tammar wallaby are known to be involved in glycolysis and
protein synethsis (Freyer et al., 2003). This implies that the
choriovitelline placenta in marsupials is an active site of metabolism
and molecular synthesis, as well as the transport of nutrients from
the maternal tissues. This is a distinct difference that is not seen
in significant shifts on other lineages, and may be due to the
formation of the placenta in the opossum from the yolk sac. In
addition, the urea cycle is a highly enriched functional annotation
cluster which directly correlates with the high levels of urea in the
Monodelphis placenta (Freyer et al., 2007). This is potentially a
Monodelphis-specific characteristic as discussed above.

The opossum placenta shows expression of genes that either do not
exist in eutherians or are not found in the term eutherian placenta
(Table X). The existence of embryonic forms of hemoglobin in fetal
tissues at term, as evidenced by the expression of HBE, HBZ, and HBO
is a feature of marsupials that is generally due to their very short
gestation periods. Embryonic hemoglobins in marsupials have a lower
affinity for oxygen and allow for a lower rate of oxygen uptake from
maternal tissues (Holland and Gooley, 1997). One novel uncharacterized
gene that shows high expression (XM\_001364916.1) is not found in
eutherians; Ensembl gene trees reveal that homologs of this gene exist
in other amniotes such as birds and turtles, but are not present in
eutherian mammals. The homolog of this gene in chicken is found to
expressed in reproductive tract (Froman et al., 2006), but its
function is unknown. WFDC2 is a member of the whey acidic protein
(WAP) family and involved in innate immunity. It is expressed in the
mammary gland of the tammar wallaby during pregnancy and early
lactation (Sharp et al., 2007), and here in abundance in the opossum
placenta.

>>>>

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="LGALS13" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{\genetreeexplanation{\textit{LGALS13}}}
\label{fig:galectins_lgals13_evolution_tree}
\end{figure}
%$
Galectins are recognized as major regulators of pregnancy in both
fetal and maternal tissues
\cite{Than.ea2012:Galectinsguardiansofeutherian}. Indeed, the
evolution of new lineage-specific galectins has occurred in both
primates and ruminants, with the genes \textit{LGALS13},
\textit{LGALS14}, and \textit{LGALS16} recognized as being
placenta-specific in \human and regulating immune responses by
inducing apoptosis of maternal T-cells
\cite{Than.ea2009:primatesubfamilyofgalectins}. Indeed,
\textit{LGALS14} has the highest expression of any galectin in the
\human and \smonkey placenta
(\cref{fig:hormone_genes,fig:galectins_lgals13_evolution_tree}), and
other primate galectins such as \textit{LGALS13} and \textit{LGALS16}
are also expressed in \human and \smonkey placenta. However, in the
closely related \bonobo, \textit{LGALS14} is not found, and
\textit{LGALS1} is the most abundant galectin. \textit{LGALS1} shows
generally high expression in all placentas analyzed. \textit{LGALS1}
is also involved in mediating maternal-fetal immune tolerance, and is
expressed in both maternal and fetal tissue in primates
\cite{Than.ea2008:Emergenceofhormonaland}. \textit{LGALS1} placental
expression was thought to be confined to eutherian mammals, as there
is evidence of adaptive evolution at this node, as well as the
appearance of estrogen response elements in eutherian cis-regulatory
regions \cite{Than.ea2008:Emergenceofhormonaland}. It was unclear as
to whether galectins such as \textit{LGALS1} were important in the
regulation of pregnancy in marsupials, as well as eutherians. Here we
show substantial expression of galectins in the opossum, especially
\textit{LGALS1} (Table 8). In addition, \textit{LGALS2} is also
expressed. The presence of galectins at the maternal-fetal interface
in \opossum implies that it was also present in the ancestral placenta
of therian mammals. Indeed, adaptive evolution of \textit{LGALS1} on
the eutherian stem is likely due to the increased invasive nature of
eutehrian placentation rather than the origination of maternal-fetal
tolerance. Although the function of \textit{LGALS1} in the marsupial
placenta needs to be confirmed, it is likely that the mediation of
maternal-fetal immune tolerance by galectins is an ancient feature of
mammalian viviparity.

@@

The placenta of \opossum is distinctly different from that of the term
placenta of eutherian mammals in that it is derived from both the
chorion and yolk sac. Other differences between the placentas of
marsupials and eutherians include lack of invasive extravillous
trophoblasts in marsupials, the retention of a permeable shell coat
for 2/3 of marsupial gestation, and the greater degree of imprinting
in placenta-expressed genes in eutherians
\cite{Renfree2010:ReviewMarsupialsplacentalmammals,Renfree.ea2013:originandevolutionof}.
In addition, the placenta of \opossum is also distinct from other
marsupials, in that it is considered one of the least specialized, and
thought be closer to the original marsupial condition
\cite{Renfree2010:ReviewMarsupialsplacentalmammals}. In \opossum, the
duration of direct connection between maternal and fetal tissues is as
short as two days in a gestational period of 13-15 days
\cite{Harder.ea1993:Gestationandplacentationin}. During this period,
the placenta is directly engaged in the exchange of urea with maternal
tissues, owing to the lack of a vesicular allantois
\cite{Freyer.ea2007:Placentalfunctionintwo}. This differs drastically
with the situation in another well-studied marsupial, the tammar
wallaby \cite{Freyer.ea2007:Placentalfunctionintwo}. Therefore, we
must interpret differences between the opossum and eutherian placentas
with caution, as highly expressed genes and functional annotation
clusters may represent didelphid or \opossum specific adaptations.

% Genes that show a significant shift on the \opossum lineage include
% functional annotation clusters that correspond to active generation of
% energy, transport, and lipid breakdown via the peroxisome (Table 10;
% Supplemental Table X). The choriovitelline placenta of both didelphids
% and the tammar wallaby are known to be involved in glycolysis and
% protein synethsis (Freyer et al., 2003). This implies that the
% choriovitelline placenta in marsupials is an active site of metabolism
% and molecular synthesis, as well as the transport of nutrients from
% the maternal tissues. This is a distinct difference that is not seen
% in significant shifts on other lineages, and may be due to the
% formation of the placenta in the opossum from the yolk sac. In
% addition, the urea cycle is a highly enriched functional annotation
% cluster which directly correlates with the high levels of urea in the
% Monodelphis placenta (Freyer et al., 2007). This is potentially a
% Monodelphis-specific characteristic as discussed above.

The opossum placenta shows expression of genes that either do not
exist in eutherians or are not found in the term eutherian placenta
(\cref{tab:top50_opossum_placenta}). The existence of embryonic forms
of hemoglobin in fetal tissues at term, as evidenced by the expression
of \textit{HBE} and \textit{HBZ} is a feature of marsupials that is
generally due to their very short gestation periods. Embryonic
hemoglobins in marsupials have a lower affinity for oxygen and allow
for a lower rate of oxygen uptake from maternal tissues
\cite{Holland.ea1997:Characterizationofembryonicglobin}. One novel
uncharacterized gene that shows high expression (LOC10001195) is not
found in eutherians; Ensembl gene trees reveal that homologs of this
gene exist in other amniotes such as birds and turtles, but are not
present in eutherian mammals. The homolog of this gene in chicken is
found to expressed in reproductive tract
\cite{Froman.ea2006:expressedsequencetaganalysis}, but its function is
unknown. \textit{WFDC2} is a member of the whey acidic protein (WAP)
family and involved in innate immunity. It is expressed in the mammary
gland of the tammar wallaby during pregnancy and early lactation
\cite{Sharp.ea2007:Molecularevolutionofmonotreme}, and here in
relative abundance in the opossum placenta.
